Numerical simulation of the Cordilleran ice sheet \\ through the last glacial cycle

### 2014-07-08 kappa.tex
Add Vancouver Island wet climate photo.
Add fig, table and biblio refs in model set-up.

% Base class and packages
\documentclass{article}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{lineno}
\usepackage{natbib}
\usepackage[onehalfspacing]{setspace}
\usepackage{physics}
\usepackage[colorlinks, citecolor=blue]{hyperref}
\usepackage[sort]{cleveref}  % needs to be loaded after hyperref
\usepackage{graphicx}
\graphicspath{{figures/}}

@@

% Alternative code using subcaption package
%\usepackage{subcaption}
%\begin{figure}
%  \centering
%  \makebox[0pt]{
%    \begin{subfigure}[t]{80mm}
%        \includegraphics{photo-glacier-crevasses}
%        \caption{Subcaption a.}
%        \label{fig:1a}
%    \end{subfigure}
%    \hspace{10mm}
%    \begin{subfigure}[t]{80mm}
%        \includegraphics{photo-glacier-fold}
%        \caption{Subcaption b.}
%        \label{fig:1b}
%    \end{subfigure}
%  }
%  \caption{Global caption.}
%  \label{fig:1}
%\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics{photo-fraser-valley}
  \caption{This dry landscape of the Fraser Valley, central British Columbia,
           exhibits climatic contrasts at play in the Cordillera.}
  \label{fig:photo-fraser-valley}
\end{figure}

@@

The thermodynamic core of the ice sheet model relies on four elementary field
equations. Firstly, ice flow is assumed \emph{incompressible}. Hence, the ice
density, $\rho$, is constant, and the
conservation of mass implies a conservation of volume, which can be expressed
in terms of the ice velocity vector, $\vv$, by
\begin{equation}
    \label{eqn:incompressibility}
    \div{\vv} = 0 \,.
\end{equation}

@@

Thirdly, after defining the strain-rate tensor
${\SRT = \frac{1}{2}(\tgrad{\vv} + \tgrad{\vv}^{\mathrm{T}})}$,
and the deviatoric stress tensor, ${\DST = \CST - p\,\IDT}$,
where $p=\frac{1}{3}\tr(\CST)$ is the hydrostatic pressure and
$\tens{\delta}$ is the identity tensor, a \emph{constitutive law} for ice based
on laboratory experiments is used to relate these two quantities. It reads:
\begin{equation}
    \label{eqn:glenslaw}
    \SRT = A\,\tau_e^{n-1}\,\DST \,.
\end{equation}
The equivalent stress, $\tau_e$, is defined by
${\tau_e}^2 = -\mathrm{II}_{\DST} = \frac{1}{2} \tr(\DST^2)$,
where $\mathrm{II}_{\DST}$ is the second invariant of the stress tensor.
The ice softness coefficient, $A$, depends on ice temperature, pressure, and
water content through an Arrhenius-type law,
\begin{equation}
    A =
    \begin{cases}
        A_c (1+f\omega)\,e^\frac{-Q_c}{RT_{pa}}
            & \text{if}\ T < T_c \,, \\
        A_w (1+f\omega)\,e^\frac{-Q_w}{RT_{pa}}
            & \text{if}\ T \ge T_c \,,
    \end{cases}
\end{equation}
where $T_{pa}$ is the pressure-adjusted temperature calculated using the
Clapeyron relation, ${T_{pa} = T - \beta p}$. The water fraction, $\omega$, is
capped at a threshold of 0.01, above which no measurements are available.

Letting ${\doteps_e}^2 = \frac{1}{2} \tr(\SRT^2)$ and $B=A^{-1/n}$, the
constitutive law (\ref{eqn:glenslaw}) can be rewritten using a sometimes more
convenient, inverted formulation,
\begin{equation}
    \label{eqn:invglenslaw}
    \DST = B\,\doteps_e^{1/n-1}\,\SRT \,.
\end{equation}
By analogy to Newtonian flow, an apparent viscosity, $\nu$, can then be defined
by $\DST = \nu\,\SRT$, thus yielding

@@

Finally, the evolution of temperature within the ice is governed by the
\emph{conservation of energy}. In PISM, the conservation of energy uses an
enthalpy formulation in order to account for thermodynamic effects associated
with internal phase changes in the glacier. This enthalpy ($H$) formulation
reads
\begin{equation}
    \label{eqn:enthalpy}
    \rho \left(\matdv{H}\right)
        = -\div{\vect{q}} + \frac{\nu\doteps_e^2}{4} \,,
\end{equation}
where $\vect{q}$ represents the heat flow and
${\frac{\nu\dot{\epsilon_e}^2}{4} = \tr(\DST\SRT)}$ is a
source term corresponding to strain heat release. In the case where ice
temperature, $T$, is below the pressure-melting point, the enthalpy can be
expressed as $H=c\,T$, while the heat flow is given by Fourier's law,
$\vect{q} = k\,\grad{T}$. Consequently, the enthalpy equation
(\ref{eqn:enthalpy}) can be rewritten using a more familiar, cold-ice,
temperature formulation,
\begin{equation}
    \label{eqn:temperature}
    \matdv{T} = \frac{k}{\rho c} \Delta T
                + \frac{\nu\doteps_e^2}{4\rho c} \,,
\end{equation}
where $c$ denotes the specific heat capacity of the ice and $k$ is its thermal
conductivity.

@@

\begin{table}
  \centering
  \caption{Constant parameters of the ice flow model.
           \todo{find a better reference than \citet{Aschwanden.etal.2012}
                 for all the physical constants. Double-check
                 \citet{Paterson.Budd.1982} and add missing references.}}
  \label{tab:iceflowparams}
  \makebox[0pt]{\begin{tabular*}{165mm}{@{\extracolsep{\fill}}rlrll}
    \hline
    Not.    & Name & Value & Unit & Source \\
    \hline

@@

PISM, the ice sheet model that we use, employ two well-documented
approximations of ice flow equations: the shallow ice approximation and the
the shallow shelf approximation. Although both approximations can be derived
from a rigorous scaling approach, we don't expand these lengthy derivations
here, assuming instead the simplifications and focusing on their effect.

@@

Projecting along the three Cartesian coordinate axis, this reads
\begin{subequations}
\label{eqn:fullstokes}
\begin{align}
    \pdv{\tau_{xx}}{x} + \pdv{\tau_{xy}}{y} + \pdv{\tau_{xz}}{z}
        &= \pdv{p}{x} \,, \\
    \pdv{\tau_{yx}}{x} + \pdv{\tau_{yy}}{y} + \pdv{\tau_{yz}}{z}
        &= \pdv{p}{y} \,, \\
    \pdv{\tau_{zx}}{x} + \pdv{\tau_{zy}}{y} + \pdv{\tau_{zz}}{z}
        &= \pdv{p}{z} - \rho g \,.
\end{align}
\end{subequations}

The shallow ice approximation (SIA) assumes that basal friction is high enough
for the vertical shear stresses $\{\tau_{xz}, \tau_{yz}\}$ to predominate over
all other components of the deviatoric stress tensor. In this framework, the
projected stress-balance can be simplified to
\begin{subequations}
\label{eqn:sia}
\begin{align}
    \label{eqn:siax}
    \pdv{\tau_{xz}}{z} &= \pdv{p}{x} \,, \\
    \label{eqn:siay}
    \pdv{\tau_{yz}}{z} &= \pdv{p}{y} \,, \\
    \label{eqn:siaz}
    0 &= \pdv{p}{z} - \rho g \,.
\end{align}
\end{subequations}

@@

The shallow shelf approximation (SSA), on the opposite of the SIA, assumes that
basal friction is low enough for ice to deform predominantly by horizontal
expansion within the entire ice column. In this context, the stress-balance can
be simplified to
\begin{subequations}
\label{eqn:ssa}
\begin{align}
    \label{eqn:ssax}
    \pdv{\tau_{xx}}{x} + \pdv{\tau_{xy}}{y} + \pdv{\tau_{xz}}{z}
        &= \pdv{p}{x} \,, \\
    \label{eqn:ssay}
    \pdv{\tau_{yx}}{x} + \pdv{\tau_{yy}}{y} + \pdv{\tau_{yz}}{z}
        &= \pdv{p}{y} \,, \\
    \label{eqn:ssaz}
    \pdv{\tau_{zz}}{z} &= \pdv{p}{z} - \rho g \,,
\end{align}
\end{subequations}
and horizontal velocities can be assumed independent of depth, so that
\begin{align}
    \pdv{v_x}{z} &= 0 \,, \\
    \pdv{v_y}{z} &= 0 \,.
\end{align}

@@

A final integration over the entire ice column yields
\begin{subequations}
\begin{align}
    \label{eqn:vssa}
    \pdv{x} \left[2\bar{\nu}h
                  \left(2\pdv{v_x}{x} + \pdv{v_y}{y}\right) \right]
        + \pdv{y} \left[\bar{\nu}h
                        \left(\pdv{v_x}{y} + \pdv{v_y}{x}\right) \right]
        + \tau_{bx} &= \rho gh \pdv{s}{x} \,, \\
    \pdv{x} \left[\bar{\nu}H
                  \left(\pdv{v_x}{y} + \pdv{v_y}{x}\right) \right]
        + \pdv{y} \left[2\bar{\nu}h
                        \left(\pdv{v_x}{x} + 2\pdv{v_y}{y}\right) \right]
        + \tau_{by} &= \rho gh \pdv{s}{y} \,,
\end{align}
\end{subequations}
where $\tau_{bx}$ and $\tau_{by}$ correspond to $x$ and $y$-components of the
basal traction, and will be determined by a sliding law.

In PISM, the SIA and SSA velocities are eventually added,
\begin{equation}
    \label{eqn:siassa}
    \vv = \vsia + \vssa \,.
\end{equation}
Because $\vsia$ is small where $\vssa$ is large and
reciprocally, we can expect the model to perform well in regions where
conditions for applicability of either the SIA or the SSA are met. However,
\cref{eqn:siassa} is a heuristic, and its validity in the zone of
transition between shear flow and longitudinal flow has not yet been
mathematically proven.

@@

Although the SIA is valid only under non-sliding conditions, the SSA requires
a sliding law as basal boundary condition, which in fact constitutes the main
control on SSA velocities and in turn locations of ice streams.

The simulations presented in this thesis use a pseudo-plastic sliding law,
\begin{equation}
    \label{eqn:pseudoplastic}
    \vect{\tau}_b = -\tau_c \frac{\vv_b}{{v_{th}}^q\,|\vv_b|^{1-q}} \,,
\end{equation}
where $\vect{\tau}_b$ is the basal traction force, $\vv_b$ is the basal
velocity, and $v_{th}$ is an arbitrary velocity threshold. In the case of
$q=0$, \cref{eqn:pseudoplastic} corresponds to a purely plastic law,
\begin{equation}
    \label{eqn:plastic}
    \vect{\tau}_b = -\tau_c \frac{\vv_b}{|\vv_b|} \,.
\end{equation}

However a non-zero exponent is chosen here in order to improve convergence. The
yield stress $\tau_c$ is related to till properties by the Mohr-Coulomb
criterion,
\begin{equation}
   \tau_c = c_0 + N\,\tan{\phi} \,.
\end{equation}

The till cohesion $c_0$ is assumed to be zero. The effective pressure on the
till, $N$, is determined by the modelled amount of water at the bed. However,
two different parametrisations are used within this thesis. In {\CCLI}, using
PISM~0.5, the effective pressure is linearly related to water content by a
simple parametrisation,
\begin{equation}
    N = \rho gh (1 - \alpha \frac{W}{W_{max}}) \,.
\end{equation}

In {\CCYC}, using PISM~0.6, the effective pressure is determined by an
empirical parametrisation based on laboratory experiments on Antarctic till,
\begin{equation}
    N = \delta \rho gh \, 10^{(e_0/C_c) (1 - (W/W_{max}))} \,.
\end{equation}

Drainage of water at the base of the ice sheet, although implemented in PISM,
is not included in simulations presented here for the sake of simplicity.
In turn, the amount of water at the bed, $W$, corresponds to the local
accumulation of basal melt-water. In varies from zero to $W_{max}$, a threshold
above which further melt exits the system.

Finally, the till friction angle $\phi$ is taken a~function of modern bed
elevation, with lowest values occurring at low elevations, thus accounting for
a weakening of till associated with the presence of marine sediments:
\begin{equation}
    \phi(x,y) =
    \begin{cases}
        \phi_0 & \text{if}\ b \le b_0 \,, \\
        \phi_0 + (\phi_1-\phi_0) \frac{b - b_0}{b_1-b_0}
                & \text{if}\ b_0 < b < b_1 \,, \\
        \phi_1 & \text{if}\ b_1 \le b \,,
    \end{cases}
\end{equation}
where $b_0=0$\,m is modern sea-level and $b_1=200$\,m is a rough average of
highest shoreline elevations in the model domain. The values of $\phi_1$ and
$\phi_2$ are chosen differently in \CCLI (10 and 30$^\circ$) and \CCYC
(15 and 45$^\circ$), but these choices are roughly equivalent in the case of a
saturated till ($W=W_{max}$).

@@

Secondly, bedrock elevation evolve in response to the ice load. The bedrock
deformation model used in this thesis describes the flexure of an elastic
lithosphere on top of an infinite half-space viscous mantle
\citep{Lingle.Clark.1985}. It can be described by a single differential
equation of the bed elevation, $b$,
\begin{equation}
    2\nu_m\,|\grad|\,\pdv{b}{t} + \rho_l g b + D\,\Delta^2 b = \sigma_{zzb} \,,
\end{equation}
where $\nu_m$ is mantle viscosity, $\rho_l$ is lithosphere density, $D$ is the
lithosphere's flexural rigidity, and $\sigma_{zzb}$ corresponds to the ice load
\citep{Bueler.etal.2007}. Here, $\Delta^2$ designs the biharmonic (Laplacian
square) operator of
linear elastic theory, while $|\grad|$ is a pseudo-differential operator
defined through the Fourier transform by \citet[Eq.~6]{Bueler.etal.2007}. In
the left-hand part of this equation, the first term accounts for mantle
relaxation, the second for point-wise isostasy, and the third for elastic
flexure. Due to high mantle viscosity, $\nu_m$, there exists a time lag between
the ice sheet growth and the isostatic bedrock response.
%$\nu=1\times10^{21}$\,Pa\,s
%$D=5\times10^{24}$\,N\,m
%$\rho_r = 3300$\,kg\,m$^{-3}$

\begin{table}
  \centering
  \caption{Constant parameters of the bedrock model.
           \todo{Find out what the bedrock density corresponds to. Track down
                 original sources in \citet{Ritz.1997}}}
  \label{tab:bedrockparams}
  \makebox[0pt]{\begin{tabular*}{165mm}{@{\extracolsep{\fill}}rlrll}
    \hline
    Not.    & Name & Value & Unit & Source \\
    \hline

@@

At the interface between the ice sheet and the atmosphere (Fig.~??),
surfice mass fluxes are computed from monthly mean surface air temperature,
$T_m$, and monthly precipitation, $P_m$, by a temperature-index model
\citep[e.g.,][]{Hock.2003}. Surface accumulation, $a_s$, equals precipitation
when temperature is below ${T_0=0\,\degC}$, and decreases to zero linearly
with temperature between $T_0$ and ${T_1=2\,\degC}$,
\begin{equation}
    a_s =
    \begin{cases}
        0       & \text{if}\ T_m \le T_s \,, \\
        P_m \frac{T_m-T_s}{T_r-T_s}
                & \text{if}\ T_s < T_m < T_r \,, \\
        P_m     & \text{if}\ T_r \le T_m \,,
    \end{cases}
\end{equation}

@@

In the simulations presented here, \cref{eqn:calovgreve} is numerically
approximated using week-long sub-intervals. For each sub-interval, surface melt
is computed from PDDs and the available snow and ice depths using an algorithm
documented in the on-line documentation of PISM \citep{PISM-authors.2014}. This
algorythms employ different melt factors for snow ($F_s$) and for ice ($F_i$),
as derived from mass-balance measurements on contemporary glaciers in the
Coast and Rocky Mountains of British Columbia \citep{Shea.etal.2009}. In
{\CCLI}, the temperature standard deviation $\sPDD$ is a constant model
parameter. In {\CCYC}, it is computed from daily temperature values from the
North American Regional Reanalysis \citep[NARR,][]{Mesinger.etal.2006}, using
a method initially developed in {\PSDV}, and further improved in {\PSDP}.
%3.04\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for snow and
%4.59\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for ice

@@

\CCLI tests the sensitivity of the model to different input present-day
climatologies, $\{T_{m0}, P_{m0}\}$, while the temperature-offset corrections,
${\Delta}T_{TS}$, uses constant values ranging from -10 to 0\,\degC. \CCYC,
on the opposite, tests model sensitivity to time-dependent temperature-offset
corrections derived from different palaeo-temperature proxy records.

@@

>>>>

% Base class and packages
\documentclass{article}
\usepackage{amsmath}
\usepackage{bm}
\usepackage[T1]{fontenc}
\usepackage{lineno}
\usepackage{natbib}
\usepackage[onehalfspacing]{setspace}
\usepackage{physics}
\usepackage[colorlinks, citecolor=blue]{hyperref}
\usepackage[sort]{cleveref}  % needs to be loaded after hyperref
\usepackage{graphicx}
\graphicspath{{figures/}}

@@

@@

\begin{figure}
  \centering
  \makebox[0pt]{
    \subgraphics{a}{photo-vancouver-island}
    \hspace{1cm}
    \subgraphics{b}{photo-fraser-valley}
  }
  \caption{Contrasted climates of the North American Cordillera.
           \textbf{(a)} Pacific Coast at Vancouver Island.
           \textbf{(b)} Fraser Valley, south-central British Columbia.}
  \label{fig:photo-fraser-valley}
\end{figure}

@@

The thermodynamic core of the ice sheet model relies on four elementary field
equations. Firstly, ice flow is assumed \emph{incompressible}. Hence, the ice
density, $\rho$, is constant, and the
conservation of mass implies a conservation of volume, which can be expressed
in terms of the ice velocity vector, $\vv$ \cref{fig:model-variables}, by
\begin{equation}
    \label{eqn:incompressibility}
    \div{\vv} = 0 \,.
\end{equation}

@@

Thirdly, after defining the strain-rate tensor
${\SRT = \frac{1}{2}(\tgrad{\vv} + \tgrad{\vv}^{\mathrm{T}})}$,
and the deviatoric stress tensor, ${\DST = \CST - p\,\IDT}$,
where $p=\frac{1}{3}\tr(\CST)$ is the hydrostatic pressure and
$\tens{\delta}$ is the identity tensor, a \emph{constitutive law} for ice
\citep{Nye.1953} based on laboratory experiments \citep{Glen.1952} is used to
relate these two quantities. It reads:
\begin{equation}
    \label{eqn:glenslaw}
    \SRT = A\,\tau_e^{n-1}\,\DST \,.
\end{equation}
The equivalent stress, $\tau_e$, is defined by
${\tau_e}^2 = -\mathrm{II}_{\DST} = \frac{1}{2} \tr(\DST^2)$,
where $\mathrm{II}_{\DST}$ is the second invariant of the stress tensor.
The ice softness coefficient, $A$, depends on ice temperature, pressure, and
water content through an Arrhenius-type law
\citep[Eqs.~63--65]{Aschwanden.etal.2012},
\begin{equation}
    A =
    \begin{cases}
        A_c (1+f\omega)\,e^\frac{-Q_c}{RT_{pa}}
            & \text{if}\ T < T_c \,, \\
        A_w (1+f\omega)\,e^\frac{-Q_w}{RT_{pa}}
            & \text{if}\ T \ge T_c \,,
    \end{cases}
\end{equation}
where $T_{pa}$ is the pressure-adjusted temperature calculated using the
Clapeyron relation, ${T_{pa} = T - \beta p}$. The water fraction, $\omega$, is
capped at a threshold of 0.01, above which no measurements are available
\citep{Lliboutry.Duval.1985}.

\todo{I am trusting PISM's manual on that. Check other papers by Duval, or
      the book by Greve (library).}

Letting ${\doteps_e}^2 = \frac{1}{2} \tr(\SRT^2)$ and $B=A^{-1/n}$, the
constitutive law (\ref{eqn:glenslaw}) can be rewritten using a sometimes more
convenient, inverted formulation,
\begin{equation}
    \label{eqn:invglenslaw}
    \DST = B\,\doteps_e^{1/n-1}\,\SRT \,.
\end{equation}
By analogy to Newtonian flow, an apparent viscosity, $\nu$, can then be defined
by $\DST = 2 \nu\,\SRT$, thus yielding

@@

Finally, the evolution of temperature within the ice is governed by the
\emph{conservation of energy}. In PISM, the conservation of energy uses an
enthalpy formulation \citep[Eqs.~20--21]{Aschwanden.etal.2012} in order to
account for thermodynamic effects associated with internal phase changes in the
glacier. This enthalpy ($H$) formulation reads
\begin{equation}
    \label{eqn:enthalpy}
    \rho \left(\matdv{H}\right)
        = -\div{\vect{q}} + \frac{\nu\doteps_e^2}{4} \,,
\end{equation}
where $\vect{q}$ represents the heat flow and
${\frac{\nu\dot{\epsilon_e}^2}{4} = \tr(\DST\SRT)}$ is a
source term corresponding to strain heat release. In the case where ice
temperature, $T$, is below the pressure-melting point, the enthalpy can be
expressed as $H=c\,T$, while the heat flow is given by Fourier's law,
$\vect{q} = k\,\grad{T}$. Consequently, the enthalpy equation
(\ref{eqn:enthalpy}) can be rewritten using a more familiar, cold-ice,
temperature formulation,
\begin{equation}
    \label{eqn:temperature}
    \matdv{T} = \frac{k}{\rho c} \Delta T
                + \frac{\nu\doteps_e^2}{4\rho c} \,,
\end{equation}
where $c$ denotes the specific heat capacity of the ice and $k$ is its thermal
conductivity. Constant parameters of the ice flow model use in this thesis are
given in \cref{tab:iceflowparams}.

@@

\begin{table}
  \centering
  \caption{Constant parameters of the ice flow model.
           \todo{find a better reference than \citet{Aschwanden.etal.2012}
                 (library) for all the physical constants. Double-check
                 \citet{Paterson.Budd.1982} and add missing references.}}
  \label{tab:iceflowparams}
  \makebox[0pt]{\begin{tabular*}{165mm}{@{\extracolsep{\fill}}rlrll}
    \hline
    Not.    & Name & Value & Unit & Source \\
    \hline

@@

PISM, the ice sheet model that we use, employs two well-documented
approximations of ice flow equations: the shallow ice approximation
    \citep[SIA,][]{Morland.Johnson.1980, Hutter.1983}
and the the shallow shelf approximation
    \citep[SSA,][]{Morland.1987, Weis.etal.1999}.
Although both approximations can be derived
from a rigorous scaling approach, we don't expand these lengthy derivations
here, assuming instead the simplifications and focusing on their effect.

@@

Projecting along the three Cartesian axes, this reads
\begin{subequations}
\label{eqn:fullstokes}
\begin{align}
    \pdv{\tau_{xx}}{x} + \pdv{\tau_{xy}}{y} + \pdv{\tau_{xz}}{z}
        &= \pdv{p}{x} \,, \\
    \pdv{\tau_{yx}}{x} + \pdv{\tau_{yy}}{y} + \pdv{\tau_{yz}}{z}
        &= \pdv{p}{y} \,, \\
    \pdv{\tau_{zx}}{x} + \pdv{\tau_{zy}}{y} + \pdv{\tau_{zz}}{z}
        &= \pdv{p}{z} - \rho g \,.
\end{align}
\end{subequations}

The \emph{shallow ice approximation} (SIA) assumes that basal friction is high
enough
for the vertical shear stresses $\{\tau_{xz}, \tau_{yz}\}$ to predominate over
all other components of the deviatoric stress tensor. In this framework, the
projected stress-balance can be simplified to
\begin{subequations}
\label{eqn:sia}
\begin{align}
    \label{eqn:siax}
    \pdv{\tau_{xz}}{z} &= \pdv{p}{x} \,, \\
    \label{eqn:siay}
    \pdv{\tau_{yz}}{z} &= \pdv{p}{y} \,, \\
    \label{eqn:siaz}
    0 &= \pdv{p}{z} - \rho g \,.
\end{align}
\end{subequations}

\todo{\citet{Morland.Johnson.1980} derived the SIA in 2D, using a stream
      function. Their equations look very different from here. Go to
      geolibrary and check the book by \citet{Hutter.1983}, if it is there.}

@@

\todo{Add a few words about the fact that $\vsia$ can be directly obtained
      after one vertical integration. Show that vertical velocities can be
      obtained from the compressibility equation.}

The \emph{shallow shelf approximation} (SSA), on the opposite of the SIA,
assumes that
basal friction is low enough for ice to deform predominantly by horizontal
expansion within the entire ice column. In this context, the stress-balance can
be simplified to \citep[Eqs.~4.10--4.12]{Weis.etal.1999}
\begin{subequations}
\label{eqn:ssa}
\begin{align}
    \label{eqn:ssax}
    \pdv{\tau_{xx}}{x} + \pdv{\tau_{xy}}{y} + \pdv{\tau_{xz}}{z}
        &= \pdv{p}{x} \,, \\
    \label{eqn:ssay}
    \pdv{\tau_{yx}}{x} + \pdv{\tau_{yy}}{y} + \pdv{\tau_{yz}}{z}
        &= \pdv{p}{y} \,, \\
    \label{eqn:ssaz}
    \pdv{\tau_{zz}}{z} &= \pdv{p}{z} - \rho g \,,
\end{align}
\end{subequations}
and horizontal velocities can be assumed independent of depth
\citep[Eq.~4.22]{Weis.etal.1999}, so that
\begin{align}
    \pdv{v_x}{z} &= 0 \,, \\
    \pdv{v_y}{z} &= 0 \,.
\end{align}

@@

A final integration over the entire ice column yields
\citep[Eqs.~7.7--7.8]{Weis.etal.1999}
\begin{subequations}
\begin{align}
    \label{eqn:vssa}
    \pdv{x} \left[2\bar{\nu}h
                  \left(2\pdv{v_x}{x} + \pdv{v_y}{y}\right) \right]
        + \pdv{y} \left[\bar{\nu}h
                        \left(\pdv{v_x}{y} + \pdv{v_y}{x}\right) \right]
        + \tau_{bx} &= \rho gh \pdv{s}{x} \,, \\
    \pdv{x} \left[\bar{\nu}H
                  \left(\pdv{v_x}{y} + \pdv{v_y}{x}\right) \right]
        + \pdv{y} \left[2\bar{\nu}h
                        \left(\pdv{v_x}{x} + 2\pdv{v_y}{y}\right) \right]
        + \tau_{by} &= \rho gh \pdv{s}{y} \,,
\end{align}
\end{subequations}
where $\tau_{bx}$ and $\tau_{by}$ correspond to $x$ and $y$-components of the
basal traction, and will be determined by a sliding law.

In PISM, the SIA and SSA velocities are eventually added
\citep[Eq.~15]{Winkelmann.etal.2011},
\begin{equation}
    \label{eqn:siassa}
    \vv = \vsia + \vssa \,.
\end{equation}
Because $\vsia$ is small where $\vssa$ is large and
reciprocally, we can expect the model to perform well in regions where
conditions for applicability of either the SIA or the SSA are met
\cref{fig:model-siassa}. However,
\cref{eqn:siassa} is a heuristic, and its validity in the zone of
transition between shear flow and longitudinal flow has not yet been
mathematically proven

@@

Although the SIA is valid only under non-sliding conditions, the SSA requires
a sliding law as basal boundary condition, which in fact constitutes the main
control on SSA velocities and, in turn, the locations of ice streams.

The simulations presented in this thesis use a pseudo-plastic sliding law,
\begin{equation}
    \label{eqn:pseudoplastic}
    \vect{\tau}_b = -\tau_c \frac{\vv_b}{{v_{th}}^q\,|\vv_b|^{1-q}} \,,
\end{equation}
where $\vect{\tau}_b$ is the basal traction force, $\vv_b$ is the basal
velocity, and $v_{th}$ is an arbitrary velocity threshold. In the case of
$q=0$, \cref{eqn:pseudoplastic} corresponds to a purely plastic law
\citep[Eq.~11]{Winkelmann.etal.2011},
\begin{equation}
    \label{eqn:plastic}
    \vect{\tau}_b = -\tau_c \frac{\vv_b}{|\vv_b|} \,.
\end{equation}

\todo{\citep[supplement]{Aschwanden.etal.2013} refer to a paper by Schoof and
      Hindmarsh (2010, doi:10.1093/qjmam/hbp025). Check that.}

However a non-zero exponent $q$ is chosen here in order to improve convergence.
The yield stress $\tau_c$ is related to till properties by the Mohr-Coulomb
criterion,
\begin{equation}
   \tau_c = c_0 + N\,\tan{\phi} \,.
\end{equation}

The till cohesion $c_0$ is assumed to be zero. The effective pressure on the
till, $N$, is determined by the modelled amount of water at the bed. However,
two different parametrisations are used within this thesis. In {\CCLI}, using
PISM~0.5, the effective pressure is linearly related to water content by a
simple parametrisation \citep[Eq.~13]{Winkelmann.etal.2011},
\begin{equation}
    N = \rho gh (1 - \alpha \frac{W}{W_{max}}) \,.
\end{equation}

In {\CCYC}, using PISM~0.6, the effective pressure is determined by an
empirical parametrisation based on laboratory experiments on Antarctic till
\citep{Tulaczyk.etal.2000, Bueler.Pelt.2014},
\begin{equation}
    N = \delta \rho gh \, 10^{(e_0/C_c) (1 - (W/W_{max}))} \,.
\end{equation}

Drainage of water at the base of the ice sheet, although implemented in PISM,
is not included in simulations presented here for the sake of simplicity.
In turn, the amount of water at the bed, $W$, corresponds to the local
accumulation of basal melt-water. In varies from zero to $W_{max}$, a threshold
above which further melt simply exits the system.

Finally, the till friction angle $\phi$ is taken a~function of modern bed
elevation, with lowest values occurring at low elevations, thus accounting for
a weakening of till associated with the presence of marine sediments
\citep[cf.][Eq.~10]{Martin.etal.2011},
\begin{equation}
    \phi(x,y) =
    \begin{cases}
        \phi_0 & \text{if}\ b \le b_0 \,, \\
        \phi_0 + (\phi_1-\phi_0) \frac{b - b_0}{b_1-b_0}
                & \text{if}\ b_0 < b < b_1 \,, \\
        \phi_1 & \text{if}\ b_1 \le b \,,
    \end{cases}
\end{equation}
where $b_0=0$\,m is modern sea-level and $b_1=200$\,m is a rough average of
highest shoreline elevations in the model domain. Parameter values for the
sliding model are given in \cref{tab:slidingparams}.

@@

Secondly, bedrock elevation evolve in response to the ice load. The bedrock
deformation model used in this thesis describes the flexure of an elastic
lithosphere on top of an infinite half-space viscous mantle
\citep{Lingle.Clark.1985}. It can be described by a single differential
equation of the bed elevation, $b$,
\begin{equation}
    2\nu_m\,|\grad|\,\pdv{b}{t} + \rho_l g b + D\,\Delta^2 b = \sigma_{zzb} \,,
\end{equation}
where $\nu_m$ is mantle viscosity, $\rho_l$ is lithosphere density, $D$ is the
lithosphere's flexural rigidity, and $\sigma_{zzb}$ corresponds to the ice load
\citep{Bueler.etal.2007}. Here, $\Delta^2$ designs the biharmonic (Laplacian
square) operator of
linear elastic theory, while $|\grad|$ is a pseudo-differential operator
defined through the Fourier transform by \citet[Eq.~6]{Bueler.etal.2007}. In
the left-hand part of this equation, the first term accounts for mantle
relaxation, the second for point-wise isostasy, and the third for elastic
flexure. Due to high mantle viscosity, $\nu_m$, there exists a time lag between
the ice sheet growth and the isostatic bedrock response. Parameter values for
bedrock response model are given in \cref{tab:bedrockparams}.

\begin{table}
  \centering
  \caption{Constant parameters of the bedrock thermal and deformation models.
           \todo{Find out what the bedrock density corresponds to. Track down
                 original sources in \citet{Ritz.1997}}}
  \label{tab:bedrockparams}
  \makebox[0pt]{\begin{tabular*}{165mm}{@{\extracolsep{\fill}}rlrll}
    \hline
    Not.    & Name & Value & Unit & Source \\
    \hline

@@

At the interface between the ice sheet and the atmosphere
(\cref{fig:model-interfaces}),
surfice mass fluxes are computed from monthly mean surface air temperature,
$T_m$, and monthly precipitation, $P_m$, by a temperature-index model
\citep[e.g.,][]{Hock.2003}. Surface accumulation, $a_s$, equals precipitation
when temperature is below ${T_0=0\,\degC}$, and decreases to zero linearly
with temperature between $T_0$ and ${T_1=2\,\degC}$,
\begin{equation}
    a_s =
    \begin{cases}
        0       & \text{if}\ T_m \le T_s \,, \\
        P_m \frac{T_m-T_s}{T_r-T_s}
                & \text{if}\ T_s < T_m < T_r \,, \\
        P_m     & \text{if}\ T_r \le T_m \,,
    \end{cases}
\end{equation}

@@

In the simulations presented here, \cref{eqn:calovgreve} is numerically
approximated using week-long sub-intervals. For each sub-interval, PDDs, which
serve as a proxy for the available energy for melt, define the amount of snow
melt through a proportionality factor $F_s$. If all snow is melted, remaining
PDDs are use to melt ice, using a proportionality factor $F_i$
(\cref{fig:model-pdd}). The melt factors $F_s$ and $F_i$
(\cref{tab:surfaceparams})
as derived from mass-balance measurements on contemporary glaciers in the
Coast and Rocky Mountains of British Columbia \citep{Shea.etal.2009}. In
{\CCLI}, the temperature standard deviation $\sPDD$ is a constant model
parameter. In {\CCYC}, it is computed from daily temperature values from the
North American Regional Reanalysis \citep[NARR,][]{Mesinger.etal.2006}, using
a method initially developed in {\PSDV}, and further improved in {\PSDP}.

@@

\CCLI\ tests the sensitivity of the model to different input present-day
climatologies, $\{T_{m0}, P_{m0}\}$, while the temperature-offset corrections,
${\Delta}T_{TS}$, uses constant values ranging from -10 to 0\,\degC. \CCYC,
on the opposite, tests model sensitivity to time-dependent temperature-offset
corrections derived from different palaeo-temperature proxy records. All data
climatic and topographic data sources used are summarised in
\cref{tab:datasources}.

@@

\todo{Add here a short section about discretization of the SIA and SSA, and
      their numerical implementation. The main subtlety, I think, lies in the
      fact that some computation need to use a ``staggered'' grid in order to
      ensure that we add SIA and SSA values corresponding to the same point in
      space. I need to check that. This section is also the place where to
      introduce geographic transformation parameters that I used throughout
      the thesis -- except in \cref{fig:map-cordillera}!}
